Race Science in Mainstream Psychology, 1960-2010

### 2025-04-15 paper/paper.qmd
proofread

Considering the five strategies identified by @FernandezPintoKnowBetterNot2017, the Pioneer Fund itself clearly exemplifies the strategy of supporting friendly research; and figures such as Shockley and Jensen exemplify the strategy of recruiting high-profile, distinguished scientists [@TuckerFundingScientificRacism2002].

In @FernandezPintoKnowBetterNot2017, the paradigm for supporting friendly research is the Tobacco Industry Research Committee, later known as the Council for Tobacco Research. @FernandezPintoKnowBetterNot2017 points out that the TIRC/CTR funded "scientific experts in major universities and research institutions" rather than marginal, fringe, or industry-employed scientists (57). The purpose was "to use 'good' research as a distraction" from the harmful effects of smoking by "fund[ing] research on heredity, infection, nutrition, hormones, nervous tension, and environmental factors" as causes of lung cancer and heart disease (57). This process of "industrial selection" for favorable research [@HolmanExperimentationIndustrialSelection2017] created the misleading impression that these non-tobacco causes were just as important, perhaps even more important, than smoking [compare @WeatherallHowBeatScience2018; @OConnorMisinformationAgeHow2019]. Similarly, per Laughlin, the Pioneer Fund was explicitly created to promote research on eugenics and race science as just as important, if not more important, than non-biological ("environmentalist") explanations of racial social inequality.

For the current project, we hypothesized that MQ and certain mainstream journals played a key role in another strategy identified by @FernandezPintoKnowBetterNot2017, the "echo chamber effect." We theorize this effect as involving three components: the "margins," the "mainstream," and the "bridge" connecting them.  @FernandezPintoKnowBetterNot2017 argues that, using sponsored journals and conferences, "the tobacco industry circumvented peer review standards in publication, without compromising its façade of research supporter" (59). These kinds of marginal venues provide a "safe space" in which favorable conceptual frameworks and empirical research (or pseudo-research) could be developed, protected from the destructive critical scrutiny they would receive in the mainstream. Bridge venues are accepted as part of the mainstream, but cite and draw on work from the margins, thereby mainstreaming these marginal ideas, that is, making them appear to be part of the legitimate body of mainstream scientific research. Where supporting friendly research is a strategy more-or-less entirely within the mainstream of the scientific community, the "echo chamber" strategy is located on the margins or fringes.

@@

The primary data for this study was the full text of all available articles published in journals that had published multiple Pioneer-funded researchers, between 1960 (the year MQ was founded) and 2010.  The end year was chosen for a combination of reasons: easy access to MQ ended at 2004, and we did not want to extend the data too far beyond that point; 2010 gave us a "tidy" 50-year window of coverage; and an end year that wasn't too recent would hopefully allow us to avoid distracting controversies about researchers who are still active today.

@@

![Disproportionate presence of Pioneer-funded authors in three focal topics.\
(A) Number of authors for three focal topics from the medium vocabulary, k= 40 model, by year. Mainstream intelligence (topic 22) shows exponential-like increase while MQ race science (topic 7) and race-and-intelligence research (topic 24) remain flat. \
(B) Pioneer-funded authors as a share (%) of all authors publishing in a given topic, by year (thin lines) and 5-year running mean (thick lines). Usually less than 15% of authors in a topic are Pioneer-funded. \
(C) Share of papers (%) with at least one Pioneer-funded author, by year (thin lines) and 5-year running mean (thick lines). Not shown is 1963 for topic 22, where Robert A. Gordon was one of two coauthors of the single publication in that topic that year. After about 1985, more than 15% of papers in topic 24 have at least one Pioneer-funded author. \
(D) Comparison of author share (panel B) to paper share (panel C). Each point is one year-topic combination. Not shown here is 1963 for topic 22. Points above the dark line $y = x$ indicate Pioneer-funded authors had a greater paper share than would be expected based on author share. Length of thin vertical lines measures "presence." \
(E) Empirical density plots of "presence," mean difference between paper share and author share for 1,000 randomly drawn sets of 13 authors. Bold vertical lines are the mean values for the Pioneer-funded authors (signed length of lines in panel D). Pioneer-funded authors are similar to randomly drawn sets for topics 7 and 22, but very different for topic 24. ](img/15_presence.png){#fig-presence}

@@

Because this was a post hoc analysis, we elected to examine "presence" only for the three focal topics for the medium vocabulary, $k=40$ model (figure @fig-focal). Figure @fig-presence shows the results of this analysis. Details of the methods used here are reported in @sec-presence.

Subfigure @fig-presence A shows that topics 7 (MQ-style race science) and 24 (race-and-intelligence) had stable numbers of authors over the scope of the study, while topic 22 (mainstream intelligence) saw an exponential-like increase. Fig. @fig-presence B indicates that Pioneer-funded authors were a relatively high share of authors in topic 24, often comprising more than 10% of all authors in this topic after 1980. In contrast, these authors were almost never more than 10% of all authors in the other two topics.

@@

However, empirical research in agntology has paid less attention to the circulation of ideas within the scientific community.  @SupranAssessingExxonMobilsGlobal2023 include several peer-reviewed journal articles in their content analysis; but compare them to intramural documents and public statements, rather than the broader scientific literature.  @FernandezPintoKnowBetterNot2017 gives three examples to illustrate the "tobacco strategy": the joint public relations-research funding campaign by the tobacco industry itself; medical ghostwriting by the pharmaceutical industry; and efforts by conservative think tanks to promote a simplistic account of the causes of the Great Recession that shifted blame away from deregulation of the finance industry.  Both the first and third cases emphasize the circulation of ideas to the public, and the third does not involve academic researchers in any significant way.  Medical ghostwriting is indeed concerning as a strategy for circulating false or misleading claims in the scientific literature. But it has not been a major case in the agnotology literature in recent years. This might reflect less concern about ghostwriting in biomedical research over time: on 2025-01-16, a Web of Science search for `medic* ghostwrit*` returned 85 total papers published during 2003-2024, with the number of publications per year peaking at 11 in both 2010 and 2011 and declining to 1-2 per year in 2023 and 2024.

@@

Schematically, a research project can be thought of as using *methods* to produce *findings* that (aim to) answer *research questions*. Each of these three elements requires various assumptions to be logically relevant to the other two  [@LonginoScienceSocialKnowledge1990; @LloydAdaptationismLogicResearch2015]. For example, a particular project might use data from a national genome bank and a genome-wide association study (GWAS; method) to develop a polygenic score estimator of educational attainment (finding) with the aim of answering the question "To what extent is cognitive ability genetic?" This project must assume that a sample from a single nation is representative of humanity in general (to be relevant to the generalized framing of the question), that "cognitive ability" is a well-defined construct and worth investigating, that educational attainment is an accurate measure of cognitive ability, that the SNPs included in the genome data capture genetically-caused variation, that analysis of the polygenic scores can recover/measure this variation, that variation can be partitioned into genetic and environmental causes [@KellerMirageSpaceNature2010], and so on.

Some of these assumptions — or the findings themselves — can support scientific racist claims, either through strict deductive entailment (the kind studied in formal logic courses) or through *implicature* [@DavisImplicature2024]. The concept of "implicature" refers to the phenomenon that language use goes beyond the strict deductive entailments — "we mean more than we say." Consider the following exchange between a teenager and their parent.

>>>>

Considering the five strategies identified by @FernandezPintoKnowBetterNot2017, the Pioneer Fund itself likely exemplifies the strategy of supporting friendly research; and figures such as Shockley and Jensen exemplify the strategy of recruiting high-profile, distinguished scientists [@TuckerFundingScientificRacism2002].

In @FernandezPintoKnowBetterNot2017, the paradigm for supporting friendly research is the Tobacco Industry Research Committee, later known as the Council for Tobacco Research. @FernandezPintoKnowBetterNot2017 points out that the TIRC/CTR funded "scientific experts in major universities and research institutions" rather than marginal, fringe, or industry-employed scientists (57). The purpose was "to use 'good' research as a distraction" from the harmful effects of smoking by "fund[ing] research on heredity, infection, nutrition, hormones, nervous tension, and environmental factors" as causes of lung cancer and heart disease (57). This process of "industrial selection" for favorable research [@HolmanExperimentationIndustrialSelection2017] created the misleading impression that these non-tobacco causes were just as important, perhaps even more important, than smoking [compare @WeatherallHowBeatScience2018; @OConnorMisinformationAgeHow2019]. Similarly, per the quote from Laughlin above, the Pioneer Fund was all-but-explicitly created to promote research on eugenics and race science as just as important, if not more important, than non-biological ("environmentalist") explanations of racial social inequality.

For the current project, we hypothesized that MQ and certain mainstream journals played a key role in another strategy identified by @FernandezPintoKnowBetterNot2017, the "echo chamber effect." We theorize this effect as involving three components: the "margins," the "mainstream," and the "bridge" connecting them.  @FernandezPintoKnowBetterNot2017 argues that, using sponsored journals and conferences, "the tobacco industry circumvented peer review standards in publication, without compromising its façade of research supporter" (59). These kinds of marginal venues provide a "safe space" in which favorable conceptual frameworks and empirical research (or pseudo-research) could be developed, protected from the destructive critical scrutiny they would receive in the mainstream [@HicksYoungsPValuePlot2022]. Bridge venues are accepted as part of the mainstream, but cite and draw on work from the margins, thereby mainstreaming these marginal ideas, that is, making them appear to be part of the legitimate body of mainstream scientific research. Where supporting friendly research is a strategy more-or-less entirely within the mainstream of the scientific community, the "echo chamber" strategy is located on the margins or fringes.

@@

The primary data for this study were the full text of all available articles published in journals that had published multiple Pioneer-funded researchers, between 1960 (the year MQ was founded) and 2010.  The end year was chosen for a combination of reasons: easy access to MQ ended at 2004, and we did not want to extend the data too far beyond that point; 2010 gave us a "tidy" 50-year window of coverage; and an end year that wasn't too recent would hopefully allow us to avoid distracting controversies about researchers who are still active today.

@@

![Disproportionate presence of Pioneer-funded authors in three focal topics.\
(A) Number of authors for three focal topics from the medium vocabulary, k= 40 model, by year. Mainstream intelligence (topic 22) shows exponential-like increase while MQ race science (topic 7) and race-and-intelligence research (topic 24) remain flat. \
(B) Pioneer-funded authors as a share (%) of all authors publishing in a given topic, by year (thin lines) and 5-year running mean (thick lines). Usually less than 15% of authors in a topic are Pioneer-funded. \
(C) Share of papers (%) with at least one Pioneer-funded author, by year (thin lines) and 5-year running mean (thick lines). Not shown is 1963 for topic 22, where Robert A. Gordon was one of two coauthors of the single publication in that topic that year. After about 1985, more than 15% of papers in topic 24 have at least one Pioneer-funded author. \
(D) Comparison of author share (panel B) to paper share (panel C). Each point is one year-topic combination. Not shown here is 1963 for topic 22. Points above the dark line $y = x$ indicate Pioneer-funded authors had a greater paper share than would be expected based on author share. Length of thin vertical lines measures "presence." \
(E) Empirical density plots of "presence," mean difference between paper share and author share for 1,000 randomly drawn sets of 13 authors. Bold vertical lines are the mean values for the Pioneer-funded authors (signed length of lines in panel D). Pioneer-funded authors are similar to randomly drawn sets for topics 7 and 22, but very different for topic 24. ](img/15_presence.png){c}

@@

Because this was a post hoc analysis, we elected to examine "presence" only for the three focal topics for the medium vocabulary, $k=40$ model (figure @fig-focal). Figure @fig-presence shows the results of this analysis. Details of the methods used here are reported in @sec-presence. Note that papers were assigned to topics in a different, less conservative way than in @fig-focal.

Subfigure @fig-presence A shows that topics 7 (MQ-style race science) and 24 (race-and-intelligence) had stable numbers of authors over the scope of the study, while topic 22 (mainstream intelligence) saw an exponential-like increase. Subfigure @fig-presence B indicates that Pioneer-funded authors were a relatively high share of authors in topic 24, often comprising more than 10% of all authors in this topic after 1980. In contrast, these authors were almost never more than 10% of all authors in the other two topics.

@@

However, empirical research in agnotology has paid less attention to the circulation of ideas within the scientific community.  @SupranAssessingExxonMobilsGlobal2023 include several peer-reviewed journal articles in their content analysis; but compare them to intramural documents and public statements, rather than the broader scientific literature.  @FernandezPintoKnowBetterNot2017 gives three examples to illustrate the "tobacco strategy": the joint public relations-research funding campaign by the tobacco industry itself; medical ghostwriting by the pharmaceutical industry; and efforts by conservative think tanks to promote a simplistic account of the causes of the Great Recession that shifted blame away from deregulation of the finance industry.  Both the first and third cases emphasize the circulation of ideas to the public, and the third does not involve academic researchers in any significant way.  Medical ghostwriting is indeed concerning as a strategy for circulating false or misleading claims in the scientific literature. But it has not been a major case in the agnotology literature in recent years. This might reflect less concern about ghostwriting in biomedical research over time: on 2025-01-16, a Web of Science search for `medic* ghostwrit*` returned 85 total papers published during 2003-2024, with the number of publications per year peaking at 11 in both 2010 and 2011 and declining to 1-2 per year in 2023 and 2024.

@@

Schematically, a research project can be thought of as using *methods* to produce *findings* that (aim to) answer *research questions*. Each of these three elements requires various assumptions to be logically relevant to the other two  [@LonginoScienceSocialKnowledge1990; @LloydAdaptationismLogicResearch2015]. For example, a particular project might use data from a national genome bank and a genome-wide association study (GWAS; method) to develop a polygenic score estimator of educational attainment (finding) with the aim of answering the question "To what extent is cognitive ability genetic?" This project must assume that a sample from a single nation is representative of humanity in general (to be relevant to the generalized framing of the question), that "cognitive ability" is a well-defined construct and worth investigating, that educational attainment is an accurate measure of cognitive ability, that the SNPs included in the genome data capture genetically-caused variation, that analysis of the polygenic scores can recover/measure this variation, that variation can be partitioned into genetic and environmental causes [@KellerMirageSpaceNature2010; @TaberyStruggleUnderstandInteraction2014], and so on.

Some of these assumptions — or the findings themselves — can support scientific racist claims, either through strict deductive entailment (the kind studied in formal logic courses) or through *implicature* [@DavisImplicature2024]. The concept of "implicature" refers to the phenomenon that language use goes beyond strict deductive entailments — "we mean more than we say." Consider the following exchange between a teenager and their parent.
